The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems

### 2011-12-16 shortRange.inhomo/alltext/pubilsh/main.tex
some minor revises

\title{The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems}
\author{Han Wang}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

\begin{abstract}
  {
    In the present paper, we develop an accurate error estimate of the
    non-bonded short-range interaction for the inhomogeneous molecular
    systems.
    The root mean square force error is proved to be decomposed into
    three additive parts: the homogeneity error, the inhomogeneity error
    and the correlation error.
    % The homogeneity error is dominant
    % in the bulk regions.
    The magnitude of the inhomogeneity error, which
    is dominant in the interfacial regions, 
    can be more than one order of magnitude larger than the homogeneity
    error.
    % This is the reason why a direct simulation may waste
    % a huge amount of computational effort.
    This is the reason why a standard simulation with fixed
    cut-off radius is either less accurate if the cut-off is too small,
    or wastes considerable computational effort if the cut-off is too large.
    Therefore, based on the error estimate, the adaptive cut-off
    and long-range force correction methods are proposed to
    boost the efficiency and accuracy of the simulation, respectively.
    The
    effectiveness of the proposed methods are demonstrated by
    molecular dynamics 
    simulations of the liquid-vapor equilibrium and the nanoscale particle
    collision. Different roles of the homogeneity error and inhomogeneity
    error are also discussed.
  }
\end{abstract}

@@

\section{Introduction}
Non-bonded short-range interaction is encountered in nearly every
molecular simulation.  A naive idea to calculate short-range
interactions is to explicitly calculate all pairwise
interactions. This results in a computational cost scaling $\mathcal
O(N^2)$ per time step, which becomes rapidly inefficient as the number
of particles $N$ grows. A better way is to introduce a cut-off
radius, outside of which all pairwise interactions are simply
neglected. In combination with the cell list and the neighbor list
algorithms \cite{frenkel02b}, the total computational cost of the
short-range interactions can be reduced to an acceptable level of
$\mathcal O(N)$.  The cut-off radius neglect all pairwise interactions
outside the cut-off radius, so it introduces an error in the interaction
calculation. Since the cut-offed system can be viewed as an approximation of
the original system, it is very important to test the
convergence of properties of interest as a function of the cut-off
radius.
% before deriving conclusions on the origianl system.

Fortunately,
in most homogeneous systems, a satisfactory convergence can be
achieved even with a very small cut-off value, for example,
$2.5\sigma$.  Moreover, it is possible to correct the systematic error
of the potential energy, the pressure and the free energy by applying
the standard long range correction (LRC)~\cite{allen87a}, which
assumes the uniformity outside the cut-off radius and integrates the
contribution of the mentioned properties to infinity. 
% The spacial inhomogenety of the system in terms of particle density,
% particle type and change distribution is very common in natrual
% system.
Instead for inhomogeneous systems, in which the density of certain type of
atom changes within the length scale of several molecules, 
and thus lack of uniformity, 
the direct molecular simulation 
by the cut-off method will encounter non-trivial
difficulties.
The simulation results are found to converge only at a very large
cut-off radius value~\cite{duque2004some, ismail2007application}, also
depend on the energy/force continuity at the cut-off and the way of
applying the long-range corrections~\cite{trokhymchuk1999computer,
  guo1997long, mecke1997molecular, janecek2006long, goujon2004monte,
  shen2007comparative}.  Therefore, it is necessary to systematically
study the precision of the short-range interaction calculation, and to
develop methods for screening out the aforementioned cut-off artifact in
the inhomogeneous systems.

@@

A promising way to quantitatively analyze the undesirable cut-off
effects is to express these artifacts in terms of the difference
between the cut-offed interaction and the exact interaction, namely
the error. In homogeneous systems, the error analysis of the cut-off
method has been well established, see~\cite{kolafa1992cutoff} for
example. Also the error estimates~\cite{hummer1995numerical,
  kolafa1992cutoff, petersen1995accuracy, deserno1998mue2,
  wang2010optimizing} for long-range Ewald family
methods~\cite{ewald1921die, hockney1988computer, deserno1998mue1,
  darden1993pme, essmann1995spm} have provided a profound
understanding of the accuracy of force calculation, and introduced
parameter tuning algorithms~\cite{limbach06a, wang2010optimizing} that
boost the efficiency of the computation with a good control of the
error. However, in inhomogeneous systems the error study is still
scarce, even for the simplest cut-off method.

In this work, we develop the error estimate of the short-range force
in inhomogeneous systems. We prove that the 
force error is comprised of a homogeneity error and an inhomogeneity
error, regardless of the particle-particle correlation outside the
cut-off radius.  In an inhomogeneous system composed by several bulk
regions, an important observation is that the inhomogeneity error can
be more than one order of magnitude larger than the homogeneity
error, and plays a dominant role in the interfacial region. An
adaptive cut-off method is hereby proposed to equally distribute the
error across the simulation region, so that a substantial
computational effort can be saved while the precision of simulation is
preserved. We also proposed a long-range force correction (LFC) method
based on the error estimate to screen out the contribution of the
inhomogeneity error, which can tremendously improve the precision of
simulation without much extra computational load. To show the
effectiveness of the error estimate and the proposed methods, two
systems are tested: the liquid-vapor equilibrium for equilibrium
properties, and the colliding nanoscale clusters for dynamical
properties. Finally, conclusions are given in section
\ref{sec:conclusion}.

@@

% In Fig. \ref{fig:tmp1}, the testing system containing $N=16000$
% standard Lennard-Jone particles is set up in a tube with periodic
% boundary condition. The liquid droplet (in the middle of the tube) is
% in equilibrium with the surrounding vapor environment. The red line in
% the Fig. \ref{fig:tmp2} denotes the density profile (on direction $x$)
% of this inhomogeneous system. The green square denotes the real RMS
% force error distribution. It is clear that the error form two peaks in
% the interfacial region, which is much higher than the relative
% constant error of the bulk liquid and the bulk vapor. The dashed and
% the dotted blue line are the estimated homogeneity and inhomogeneity
% errors, and the solid blue line is the sum. In the bulk region, the
% homogeneity error dominates while in the interfacial region, the
% inhomogeneity error dominates. The esitmated error matches perfactly
% with the real error at the interfacial region, but presents
% overestimation in the bulk regions. The overestimation probably stems
% from the neglection of the correlation error in \eqref{eqn:tmp5}.

@@

In this paper, we developed the force error estimate for non-bonded
short-range interaction in the inhomogeneous molecular system.  The
RMS force error can be expressed by a summation of a homogeneity
error, an inhomogeneity error and a correlation error. In the present
paper, we only considered the cases, in which the correlation error
does not dominate. In an inhomogeneous system, the basic observation
of the error distribution was that the homogeneity error dominates the
bulk material region, whereas the inhomogeneity error dominates the
interfacial region where the density changes within the length scale
of several molecules. The inhomogeneity error can be more than one
order of magnitude larger than the homogeneity error.  Two methods
were proposed to improve the efficiency and accuracy of the
simulation. The adaptive cut-off method uses smaller cut-off radius
for the low-error bulk region and larger cut-off radius for the
high-error interfacial region, so that the RMS force error is
uniformly distributed over the simulation region. The long-range force
correction method corrects the force computation by the mean error
force, so that the inhomogeneity error is removed from the system.

>>>>

\title{The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems}
\author{Han Wang}
\email{han.wang@fu-berlin.de}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

\begin{abstract}
  {
    In the present paper, we develop an accurate error estimate of the
    non-bonded short-range interactions for the inhomogeneous molecular
    systems.
    The root mean square force error is proved to be decomposed into
    three additive parts: the homogeneity error, the inhomogeneity error
    and the correlation error.
    % The homogeneity error is dominant
    % in the bulk regions.
    The magnitude of the inhomogeneity error, which
    is dominant in the interfacial regions, 
    can be more than one order of magnitude larger than the homogeneity
    error.
    % This is the reason why a direct simulation may waste
    % a huge amount of computational effort.
    This is the reason why a standard simulation with fixed
    cut-off radius is either less accurate if the cut-off is too small,
    or wastes considerable computational effort if the cut-off is too large.
    Therefore, based on the error estimate, the adaptive cut-off
    and long-range force correction methods are proposed to
    boost the efficiency and accuracy of the simulation, respectively.
    The
    effectiveness of the proposed methods are demonstrated by
    molecular dynamics 
    simulations of the liquid-vapor equilibrium and the nanoscale particle
    collision. Different roles of the homogeneity error and inhomogeneity
    error are also discussed.
  }
\end{abstract}

@@

\section{Introduction}
Non-bonded short-range interactions are encountered in nearly every
molecular simulation.  A naive idea to calculate short-range
interactions is to explicitly calculate all pairwise
interactions. This results in a computational cost scaling $\mathcal
O(N^2)$ per time step, which becomes rapidly inefficient as the number
of particles $N$ grows. A better way is to introduce a cut-off
radius, outside of which all pairwise interactions are simply
neglected. In combination with the cell list and the neighbor list
algorithms \cite{frenkel02b}, the total computational cost of the
short-range interactions can be reduced to an acceptable level of
$\mathcal O(N)$.  The cut-off radius neglects all pairwise interactions
outside the cut-off radius, so it introduces an error in the interaction
calculation. Since the cut-offed system can be viewed as an approximation of
the original system, it is very important to test the
convergence of properties of interest as a function of the cut-off
radius.
% before deriving conclusions on the origianl system.

Fortunately,
in most homogeneous systems, a satisfactory convergence can be
achieved even with a very small cut-off value, for example,
$2.5\sigma$ for the standard Lennard-Jones 6-12 interaction.
Moreover, it is possible to correct the systematic error
of the potential energy, the pressure and the free energy by applying
the standard long range correction (LRC)~\cite{allen87a}, which
assumes the uniformity outside the cut-off radius and integrates the
contribution of the mentioned properties to infinity. 
% The spacial inhomogeneity of the system in terms of particle density,
% particle type and change distribution is very common in natrual
% system.
Instead for inhomogeneous systems, in which the density of a certain type of
atom may change within the length scale of a few molecules, 
and thus lack of uniformity, 
the standard molecular simulation 
by the cut-off method will encounter non-trivial
difficulties.
The simulation results are found to converge only at a very large
cut-off radius value~\cite{duque2004some, ismail2007application}, also
depend on the energy/force continuity at the cut-off and the way of
applying the long-range corrections~\cite{trokhymchuk1999computer,
  guo1997long, mecke1997molecular, janecek2006long, goujon2004monte,
  shen2007comparative}.  Therefore, it is necessary to systematically
study the precision of the short-range interaction calculation, and to
develop methods for screening out the aforementioned cut-off artifact in
the inhomogeneous systems.

@@

A promising way to quantitatively analyze the undesirable cut-off
effects is to express these artifacts in terms of the difference
between the cut-offed interaction and the exact interaction, namely
the error. In homogeneous systems, the error analysis of the cut-off
method has been well established, see~\cite{kolafa1992cutoff, wang2011} for
example. Also the error estimates~\cite{hummer1995numerical,
  kolafa1992cutoff, petersen1995accuracy, deserno1998mue2,
  wang2010optimizing} for long-range Ewald family
methods~\cite{ewald1921die, hockney1988computer, deserno1998mue1,
  darden1993pme, essmann1995spm} have provided a profound
understanding of the accuracy of force calculation, and introduced
parameter tuning algorithms~\cite{limbach06a, wang2010optimizing} that
boost the efficiency of the computation with a good control of the
error. However, in inhomogeneous systems the error study is still
scarce, even for the simplest cut-off method.

In this work, we develop the error estimate of the short-range force
in inhomogeneous systems. We prove that the 
force error is comprised of a homogeneity error and an inhomogeneity
error, regardless of the particle-particle correlation outside the
cut-off radius.  In an inhomogeneous system composed by several bulk
regions connected by interfacial regions where the density changes
rapidly, an important observation is that the
inhomogeneity error plays a dominant role in the interfacial regions,
and can be more than one order of magnitude larger than
the error in the bulk regions. An
adaptive cut-off method is hereby proposed to equally distribute the
error across the simulation region, so that a considerable
computational effort can be saved in the bulk regions
while the precision of simulation is
preserved. We also proposed a long-range force correction (LFC) method
to screen out the contribution of the
inhomogeneity error, therefore,  the precision of
simulation is improved without much extra computational load. To show the
effectiveness of the error estimate and the proposed methods, two
systems are tested: the liquid-vapor equilibrium for equilibrium
properties, and the colliding nanoscale clusters for dynamical
properties. Finally, conclusions are given in section
\ref{sec:conclusion}.

@@

% In Fig. \ref{fig:tmp1}, the testing system containing $N=16000$
% standard Lennard-Jones particles is set up in a tube with periodic
% boundary condition. The liquid droplet (in the middle of the tube) is
% in equilibrium with the surrounding vapor environment. The red line in
% the Fig. \ref{fig:tmp2} denotes the density profile (on direction $x$)
% of this inhomogeneous system. The green square denotes the real RMS
% force error distribution. It is clear that the error form two peaks in
% the interfacial region, which is much higher than the relative
% constant error of the bulk liquid and the bulk vapor. The dashed and
% the dotted blue line are the estimated homogeneity and inhomogeneity
% errors, and the solid blue line is the sum. In the bulk region, the
% homogeneity error dominates while in the interfacial region, the
% inhomogeneity error dominates. The esitmated error matches perfactly
% with the real error at the interfacial region, but presents
% overestimation in the bulk regions. The overestimation probably stems
% from the neglection of the correlation error in \eqref{eqn:tmp5}.

@@

In this paper, we developed the force error estimate for non-bonded
short-range interactions in the inhomogeneous molecular systems. The
RMS force error can be expressed by a summation of a homogeneity
error, an inhomogeneity error and a correlation error. In the present
paper, we only considered the cases, in which the correlation error
does not dominate. In an inhomogeneous system, the basic observation
of the error distribution was that the homogeneity error dominates the
bulk material regions, whereas the inhomogeneity error dominates the
interfacial regions where the density changes within the length scale
of a few molecules. The inhomogeneity error can be more than one
order of magnitude larger than the homogeneity error.  Two methods
were proposed to improve the efficiency and accuracy of the
simulation. The adaptive cut-off method uses smaller cut-off radius
for the low-error bulk regions and larger cut-off radius for the
high-error interfacial regions, so that the RMS force error is
uniformly distributed over the simulation region. The long-range force
correction method corrects the force computation by the mean error
force, so that the inhomogeneity error is removed from the system.
